---
title: "`SpatialData.plot`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.plot')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 1 
    toc_float: true
vignette: |
  %\VignetteIndexEntry{SpatialData.plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: >
  `SpatialData.plot` provides a visualization suit for `SpatialData` objects 
  in `ggplot2`-style, with support for multi-scale `images` and `labels`, 
  `points` and `shapes`, and layering of elements into composite plots. 
  For the latter, POINT, POLYGON, and MULTIPOLYGON geometries are supported. 
  For all but `images`, there is support to pass information from annotation 
  `tables` as aesthetics (e.g., `colData` or `assay` data). For `images`, 
  auto-contrasting and multi-channel color merging are supported. 
  Each layer returns a standard `ggplot` object, allowing for customization 
  (e.g., using `ggnewscale` for layer-specific fill/color scales), 
  or multi-panel arrangement (e.g., using `patchwork`).
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

# Installation {-}

`SpatialData.plot` can be installed as follows:

```{r inst, eval=FALSE}
if (require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("SpatialData.plot")
```

For the purpose of this vignette, we require the following dependencies:

```{r load-deps}
library(ggplot2)
library(patchwork)
library(ggnewscale) 
library(spatialdataR)
library(SpatialData.plot)
```

# Introduction

The `SpatialData.plot` package contains a set of plotting functions for 
spatial omics data stored as [SpatialData](https://spatialdata.scverse.org/en/latest/index.html) 
`.zarr` files that follow [OME-NGFF specs](https://ngff.openmicroscopy.org/latest/#image-layout).
Each `SpatialData` object is composed of five layers: images, labels, shapes, 
points, and tables. Each layer may contain an arbitrary number of elements.
Images and labels are represented as `ZarrArray`s (`r BiocStyle::Biocpkg("Rarr")`). 
Points and shapes are represented as `r BiocStyle::CRANpkg("arrow")` objects linked 
to an on-disk *.parquet* file. As such, all data are represented out of memory.
Element annotation as well as cross-layer summarizations (e.g., count matrices)
are represented as `r BiocStyle::Biocpkg("SingleCellExperiment")` as tables.

**We here demonstrate package functionality using a trivial example dataset.
For more realistic technology-focused examples, we refer readers to the 
companion [demo website](https://helenalc.github.io/SpatialData.demo).**

```{r read-data}
x <- file.path("extdata", "blobs.zarr")
x <- system.file(x, package="spatialdataR")
(x <- readSpatialData(x))
```

# Visualization

`SpatialData.plot` provides 5 user-facing plotting functions:

- `plotSpatialData()` renders a base (blank) `ggplot` with simple aesthetics
and, importantly, a fixed axial ratio as to not distort spatial coordinates.
- `plotImage/Label/Point/Shape()` render specific `SpatialData` elements;
each requires specification of the element index/name `i`, and coordinate 
system `j` (by default, the first available element/system will be rendered);
different functions further accept different arguments to control aesthetics.

## Images

`Image/LabelArray`s are linked to potentially multiscale .zarr stores.
Their show method includes the scales available for a given element:

```{r ms-show}
image(x, "blobs_image")
image(x, "blobs_multiscale_image")
```

Internally, multiscale `ImageArray`s are stored as a list of `ZarrArray`, e.g.:

```{r ms-dims}
i <- image(x, "blobs_multiscale_image")
vapply(data(i, k=NULL), dim, numeric(3))
```

To retrieve a specific scale's `ZarrArray`, we can use `data(., k)`,
where `k` specifies the target scale. This also works for plotting:

```{r ms-plot}
wrap_plots(nrow=1, lapply(seq(3), \(.) 
    plotSpatialData() + plotImage(x, i=2, k=.)))
```

## Labels

Like `image`s, `label`s are represented as `ZarrArray`. Array values, however,
are integers that correspond to different instances, typically linked to a 
`table` of annotations (e.g., expression values, observation metadata, etc.).

```{r plotLabel, fig.width=9, fig.height=3.5}
i <- "blobs_labels"
t <- getTable(x, i)
t$id <- sample(letters, ncol(t))
table(x) <- t

p <- plotSpatialData()
pal_d <- hcl.colors(10, "Spectral")
pal_c <- hcl.colors(9, "Inferno")[-9]

a <- p + plotLabel(x, i, pal="grey")                   # binary
b <- p + plotLabel(x, i, c="id", pal=pal_d)            # metadata
c <- p + plotLabel(x, i, c="channel_1_sum", pal=pal_c) # assay

(a | b | c) + 
    plot_layout(guides="collect") & 
    theme(legend.position="bottom")
```

## Points

Elements of the `points` layer are rendered using `geom_point()`, 
and accepts both discrete and continuous color specifications. 

```{r plotPoint, fig.width=10, fig.height=3}
i <- "blobs_points"
a <- p + plotPoint(x, i)
b <- p + plotPoint(x, i, col="genes")       # discrete
c <- p + plotPoint(x, i, col="instance_id") # continuous
(a | b | c) 
```

## Shapes

For demonstration, we render each `shape` element present in the object, 
which includes all possible types: circles and (multi)polygons.

```{r plotShape, fig.width=8, fig.height=3}
p <- plotSpatialData()
a <- p +
    ggtitle("polygons") +
    plotShape(x, "blobs_polygons")
b <- p +
    ggtitle("multipolygons") +
    plotShape(x, "blobs_multipolygons")
c <- p +
    ggtitle("circles") +
    plotShape(x, "blobs_circles")
(a | b | c)
```

## Layering

Several `SpatialData` elements (images, labels, points, shapes) can be layered 
on-top of each other in `ggplot`-style using the `+`-operator. The return value 
of each such call remains a `ggplot` object, which accepts additional aesthetics.
Note that specifying layer-specific fill/color requires setting up a new scale
with `ggnewscale::new_scale_fill/color()`; an example is as follows:

```{r blobs-plot, fig.width=8, fig.height=5}
p <- plotSpatialData()
# joint
all <- p +
    plotImage(x) +
    plotLabel(x, a=1/3) +
    plotShape(x, 1) +
    plotShape(x, 3) +
    new_scale_color() +
    plotPoint(x, col="genes") +
    ggtitle("layered")
# split
one <- list(
    p + plotImage(x) + ggtitle("image"),
    p + plotLabel(x) + ggtitle("labels"),
    p + plotShape(x, 1) + ggtitle("circles"),
    p + plotShape(x, 3) + ggtitle("polygons"),
    p + plotPoint(x, col="genes") + ggtitle("points"))
wrap_plots(c(list(all), one), nrow=2)
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
