---
title: "`SpatialData.plot`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.plot')`"
author: 
  - name: Helena Lucia Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Vince Carey
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData.plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(ggplot2)
library(patchwork)
library(ggnewscale)
library(SpatialData)
library(SpatialData.plot)
library(SingleCellExperiment)
```

# Introduction

The `SpatialData` package contains a set of reader and plotting functions for 
spatial omics data stored as [SpatialData](https://spatialdata.scverse.org/en/latest/index.html) 
`.zarr` files that follow [OME-NGFF specs](https://ngff.openmicroscopy.org/latest/#image-layout).

Each `SpatialData` object is composed of five layers: images, labels, shapes, 
points, and tables. Each layer may contain an arbitrary number of elements.

Images and labels are represented as `ZarrArray`s (`r BiocStyle::Biocpkg("Rarr")`). 
Points and shapes are represented as `r BiocStyle::CRANpkg("arrow")` objects linked 
to an on-disk *.parquet* file. As such, all data are represented out of memory.

Element annotation as well as cross-layer summarizations (e.g., count matrices)
are represented as `r BiocStyle::Biocpkg("SingleCellExperiment")` as tables.

```{r read-blobs}
x <- file.path("extdata", "blobs.zarr")
x <- system.file(x, package="SpatialData")
(x <- readSpatialData(x, anndataR=FALSE))
```

# Visualization

### Images

`Image/LabelArray`s are linked to potentially multiscale .zarr stores.
Their show method includes the scales available for a given element:

```{r ms-show}
image(x, "blobs_image")
image(x, "blobs_multiscale_image")
```

Internally, multiscale `ImageArray`s are stored as a list of `ZarrArray`, e.g.:

```{r ms-dims}
i <- image(x, "blobs_multiscale_image")
vapply(i@data, dim, numeric(3))
```

To retrieve a specific scale's `ZarrArray`, we can use `data(., k)`,
where `k` specifies the target scale. This also works for plotting:

```{r ms-plot}
wrap_plots(nrow=1, lapply(seq(3), \(.) 
    plotSpatialData() + plotImage(x, i=2, k=.)))
```

### Labels

```{r plotLabel, fig.width=8, fig.height=3.5}
i <- "blobs_labels"
p <- plotSpatialData()
pal_d <- hcl.colors(10, "Spectral")
pal_c <- hcl.colors(9, "Inferno")[-9]
a <- p + plotLabel(x, i) # simple binary image
b <- p + plotLabel(x, i, "instance_id", pal=pal_d) # 'colData'
c <- p + plotLabel(x, i, "channel_1_sum", pal=pal_c) # 'assay'
(a | b | c) + 
    plot_layout(guides="collect") & 
    theme(legend.position="bottom")
```


### Points

```{r plotPoint, fig.width=8, fig.height=2.5}
i <- "blobs_points"
p <- plotSpatialData()
# mock up a 'table'
f <- list(
  numbers=\(n) runif(n),
  letters=\(n) sample(letters, n, TRUE))
y <- setTable(x, i, f)
# demo. viz. capabilities
a <- p + plotPoint(y, i) # simple dots
b <- p + plotPoint(y, i, "letters") # discrete coloring
c <- p + plotPoint(y, i, "numbers") # continuous coloring
a | b | c
```

### Shapes

```{r plotShape, fig.width=6, fig.height=2.5}
p <- plotSpatialData()
a <- p +
  ggtitle("polygons") +
  plotShape(x, "blobs_polygons")
b <- p +
  ggtitle("multipolygons") +
  plotShape(x, "blobs_multipolygons")
c <- p +
  ggtitle("circles") +
  plotShape(x, "blobs_circles")
wrap_plots(a, b, c)
```

### Layering

```{r blobs-plot, fig.width=8, fig.height=5}
p <- plotSpatialData()
# joint
all <- p +
    plotImage(x) +
    plotLabel(x, a=1/3) +
    plotShape(x, 1) +
    plotShape(x, 3) +
    new_scale_color() +
    plotPoint(x, c="genes") +
    ggtitle("layered")
# split
one <- list(
    p + plotImage(x) + ggtitle("image"),
    p + plotLabel(x) + ggtitle("labels"),
    p + plotShape(x, 1) + ggtitle("circles"),
    p + plotShape(x, 3) + ggtitle("polygons"),
    p + plotPoint(x, c="genes") + ggtitle("points"))
wrap_plots(c(list(all), one), nrow=2)
```

# Examples

## MERFISH

In this example data, we do not have a `label` for the `shape` polygons. 
Such labels could be morphological regions annotated by pathologists.

```{r merfish-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="merfish.zarr.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa, anndataR=FALSE))
```

There are only `r length(SpatialData::shape(x, "cells"))` cells, but 
`r format(length(SpatialData::point(x, "single_molecule")), big.mark=",")` 
molecules, so that we downsample a random subset of 2,000 for visualization:

```{r merfish-plot}
# downsample 2,000 points
n <- length(p <- point(x))
q <- p[sample(n, 1e3)]
(point(x, "2k") <- q)
# layered visualization
plotSpatialData() +
    plotImage(x) +
    plotPoint(x, i="2k", c="cell_type", s=0.2) +
    new_scale_color() +
    plotShape(x, i="anatomical") +
    scale_color_manual(values=hcl.colors(6, "Spectral")) 
```

```{r merfish-box}
# bounding-box query
qu <- list(xmin=1800, xmax=2400, ymin=5000, ymax=5400)
bb <- geom_rect(do.call(aes, qu), data.frame(qu), col="yellow", fill=NA)
y <- SpatialData(images=list("."=do.call(query, c(list(x=image(x)), qu))))
plotSpatialData() + plotImage(x) + bb | plotSpatialData() + plotImage(y)
```

## VisiumHD

Mouse intestine, 1GB; 4 image resolutions and 3 shapes at 2, 8, and 16 $\mu$m.

```{r visiumhd-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="visium_hd_3.0.0_io.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa, images=4, shapes=3, tables=FALSE))
```

```{r visiumhd-plot}
qu <- list(xmin=100, xmax=300, ymin=200, ymax=400)
bb <- geom_rect(do.call(aes, qu), data.frame(qu), col="black", fill=NA)
y <- SpatialData(images=list("."=do.call(query, c(list(x=image(x)), qu))))
plotSpatialData() + plotImage(x) + bb | plotSpatialData() + plotImage(y)
```

## MibiTOF

Colorectal carcinoma, 25 MB; no shapes, no points.

```{r mibitof-read}
dir.create(td <- tempfile())
pa <- unzip_spd_demo(
    zipname="mibitof.zip", 
    dest=td, source="biocOSN")
(x <- readSpatialData(pa, anndataR=FALSE))
```

```{r mibitof-plot, fig.width=10, fig.height=3}
pal <- hcl.colors(8, "Spectral")
wrap_plots(nrow=1, lapply(seq(3), \(.)
    plotSpatialData() + plotImage(x, .) + 
    plotLabel(x, ., "Cluster", pal=pal))) +
    plot_layout(guides="collect")
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
